### 2023-12-07 docs/example_paper.tex
add mlsys2024 example

%%%%%%%% mlsys 2024 EXAMPLE LATEX SUBMISSION FILE %%%%%%%%%%%%%%%%%

\documentclass{article}

% Recommended, but optional, packages for figures and better typesetting:
\usepackage{microtype}
\usepackage{graphicx}
\usepackage{subfigure}
\usepackage{booktabs} % for professional tables

% hyperref makes hyperlinks in the resulting PDF.
% If your build breaks (sometimes temporarily if a hyperlink spans a page)
% please comment out the following usepackage line and replace
% \usepackage{mlsys2024} with \usepackage[nohyperref]{mlsys2024} above.
\usepackage{hyperref}

% Attempt to make hyperref and algorithmic work together better:
\newcommand{\theHalgorithm}{\arabic{algorithm}}

% Use the following line for the initial blind version submitted for review:
% \usepackage{mlsys2024}

% If accepted, instead use the following line for the camera-ready submission:
\usepackage[accepted]{mlsys2024}

% The \mlsystitle you define below is probably too long as a header.
% Therefore, a short form for the running title is supplied here:
\mlsystitlerunning{Step-by-Step: Training IMU-based Gestures with Live Feedback}

\begin{document}

\twocolumn[
\mlsystitle{Step-by-Step: Training IMU-based Gestures with Live Feedback}

% It is OKAY to include author information, even for blind
% submissions: the style file will automatically remove it for you
% unless you've provided the [accepted] option to the mlsys2024
% package.

% List of affiliations: The first argument should be a (short)
% identifier you will use later to specify author affiliations
% Academic affiliations should list Department, University, City, Region, Country
% Industry affiliations should list Company, City, Region, Country

% You can specify symbols, otherwise they are numbered in order.
% Ideally, you should not use this facility. Affiliations will be numbered
% in order of appearance and this is the preferred way.
\mlsyssetsymbol{equal}{*}

\begin{mlsysauthorlist}
\mlsysauthor{Michael Schnebly}{h}
\end{mlsysauthorlist}

\mlsysaffiliation{h}{School of Engineering and Applied Sciences, Harvard University, Cambridge, MA, USA}

\mlsyscorrespondingauthor{Michael Schnebly}{michael\_schnebly@g.harvard.edu}

% You may provide any keywords that you
% find helpful for describing your paper; these are used to populate
% the "keywords" metadata in the PDF but will not be shown in the document
\mlsyskeywords{Machine Learning, MLSys}

\vskip 0.3in

\begin{abstract}
Recognizing user-defined gestures in IMU data unlocks new forms of creativity and accessibility in human-computer interaction. However, training gesture recognition models is a difficult task that requires a deep understanding of machine learning. We present Step-by-Step, a software tool that allows users to train gesture recognition models with live visual and auditory feedback. Step-by-Step is designed to be accessible to users with no machine learning experience, while providing a powerful codebase for advanced users. This work makes the following contributions: ...
% (1) a novel interface for training gesture recognition models with live feedback, (2) a novel interface for visualizing IMU data, and (3) a novel interface for visualizing gesture recognition models.
\end{abstract}
]

% this must go after the closing bracket ] following \twocolumn[ ...

% This command actually creates the footnote in the first column
% listing the affiliations and the copyright notice.
% The command takes one argument, which is text to display at the start of the footnote.
% The \mlsysEqualContribution command is standard text for equal contribution.
% Remove it (just {}) if you do not need this facility.

\printAffiliationsAndNotice{}  % leave blank if no need to mention equal contribution
% \printAffiliationsAndNotice{\mlsysEqualContribution} % otherwise use the standard text.

\section{Electronic Submission}
\label{submission}

Submission to MLSys 2024 will be entirely electronic, via a web site
(not email). Information about the submission process and \LaTeX\ templates
are available on the conference web site at:
\begin{center}
\textbf{\texttt{https://www.mlsys.org/}}
\end{center}

The guidelines below will be enforced for initial submissions and
camera-ready copies. Here is a brief summary:
\begin{itemize}
\item Submissions must be in PDF\@.
\item The maximum paper length is \textbf{10 pages} excluding references and appendices.
    (pages 11 and on must contain only references).
\item \textbf{Do not include author information or acknowledgements} in your
    initial submission.
\item Your paper should be in \textbf{10 point Times font}.
\item Make sure your PDF file only uses Type-1 fonts.
\item Place figure captions \emph{under} the figure (and omit titles from inside
    the graphic file itself). Place table captions \emph{over} the table.
\item References must include page numbers whenever possible and be as complete
    as possible. Place multiple citations in chronological order.
\item Do not alter the style template; in particular, do not compress the paper
    format by reducing the vertical spaces.
\item Keep your abstract brief and self-contained, one paragraph and roughly
    4--6 sentences. Gross violations will require correction at the
    camera-ready phase. The title should have content words capitalized.
\item Submissions that significantly diverge from the specified format 
    will be rejected without review.
\end{itemize}

\subsection{Submitting Papers}

\textbf{Paper Deadline:} The deadline for paper submission that is
advertised on the conference website is strict. If your full,
anonymized, submission does not reach us on time, it will not be
considered for publication. There is no separate abstract submission.

\textbf{Anonymous Submission:} MLSys uses double-blind review: no identifying
author information may appear on the title page or in the paper
itself. Section~\ref{author info} gives further details.

\textbf{Simultaneous Submission:} MLSys will not accept any paper which,
at the time of submission, is under review for another conference or
has already been published. This policy also applies to papers that
overlap substantially in technical content with conference papers
under review or previously published. MLSys submissions must not be
submitted to other conferences during MLSys's review period. Authors
may submit to MLSys substantially different versions of journal papers
that are currently under review by the journal, but not yet accepted
at the time of submission. Informal publications, such as technical
reports or papers in workshop proceedings which do not appear in
print, do not fall under these restrictions.

\medskip

Authors must provide their manuscripts in \textbf{PDF} format.
Furthermore, please make sure that files contain only embedded Type-1 fonts
(e.g.,~using the program \texttt{pdffonts} in linux or using
File/DocumentProperties/Fonts in Acrobat). Other fonts (like Type-3)
might come from graphics files imported into the document.

Authors using \textbf{Word} must convert their document to PDF\@. Most
of the latest versions of Word have the facility to do this
automatically. Submissions will not be accepted in Word format or any
format other than PDF\@.

Those who use \textbf{\LaTeX} should avoid including Type-3 fonts.
Those using \texttt{latex} and \texttt{dvips} may need the following
two commands:

{\footnotesize
\begin{verbatim}
dvips -Ppdf -tletter -G0 -o paper.ps paper.dvi
ps2pdf paper.ps
\end{verbatim}}
It is a zero following the ``-G'', which tells dvips to use
the config.pdf file. Newer \TeX\ distributions don't always need this
option.

Using \texttt{pdflatex} rather than \texttt{latex}, often gives better
results. This program avoids the Type-3 font problem, and supports more
advanced features in the \texttt{microtype} package.

\textbf{Graphics files} should be a reasonable size, and included from
an appropriate format. Use vector formats (.eps/.pdf) for plots,
lossless bitmap formats (.png) for raster graphics with sharp lines, and
jpeg for photo-like images.

The style file uses the \texttt{hyperref} package to make clickable
links in documents. If this causes problems for you, add
\texttt{nohyperref} as one of the options to the \texttt{mlsys2024}
usepackage statement.

\subsection{Submitting Final Camera-Ready Copy}

The final versions of papers accepted for publication should follow the
same format and naming convention as initial submissions, except that
author information (names and affiliations) should be given. See
Section~\ref{final author} for formatting instructions.

The footnote, ``Preliminary work. Under review by the MLSys. Do not distribute.'' 
must be modified to ``\textit{Proceedings of the
$\mathit{5}^{th}$ MLSys Conference},
Santa Clara, CA, USA, 2024.
Copyright 2024 by the author(s).''

For those using the \textbf{\LaTeX} style file, this change (and others) is
handled automatically by simply changing
$\mathtt{\backslash usepackage\{mlsys2024\}}$ to
$$\mathtt{\backslash usepackage[accepted]\{mlsys2024\}}$$
Authors using \textbf{Word} must edit the
footnote on the first page of the document themselves.

Camera-ready copies should have the title of the paper as running head
on each page except the first one. The running title consists of a
single line centered above a horizontal rule which is $1$~point thick.
The running head should be centered, bold and in $9$~point type. The
rule should be $10$~points above the main text. For those using the
\textbf{\LaTeX} style file, the original title is automatically set as running
head using the \texttt{fancyhdr} package which is included in the MLSys
2024 style file package. In case that the original title exceeds the
size restrictions, a shorter form can be supplied by using

\verb|\mlsystitlerunning{...}|

just before $\mathtt{\backslash begin\{document\}}$.
Authors using \textbf{Word} must edit the header of the document themselves.

\section{Format of the Paper}

All submissions must follow the specified format.

\subsection{Length and Dimensions}

Papers must not exceed ten (10) pages, including all figures, tables,
but excluding acknowledgements, references and appendices.
Acknowledgements should be limited to grants and people who contributed to the paper.
Any submission that exceeds
this page limit, or that diverges significantly from the specified format,
will be rejected without review.

The text of the paper should be formatted in two columns, with an
overall width of 6.75~inches, height of 9.0~inches, and 0.25~inches
between the columns. The left margin should be 0.75~inches and the top
margin 1.0~inch (2.54~cm). The right and bottom margins will depend on
whether you print on US letter or A4 paper, but all final versions
must be produced for US letter size.

The paper body should be set in 10~point type with a vertical spacing
of 11~points. Please use Times typeface throughout the text.

\subsection{Title}

The paper title should be set in 14~point bold type and centered
between two horizontal rules that are 1~point thick, with 1.0~inch
between the top rule and the top edge of the page. Capitalize the
first letter of content words and put the rest of the title in lower
case.

\subsection{Author Information for Submission}
\label{author info}

MLSys uses double-blind review, so author information must not appear. If
you are using \LaTeX\/ and the \texttt{mlsys2024.sty} file, use
\verb+\mlsysauthor{...}+ to specify authors and \verb+\mlsysaffiliation{...}+ to specify affiliations. (Read the TeX code used to produce this document for an example usage.) The author information
will not be printed unless \texttt{accepted} is passed as an argument to the
style file.
Submissions that include the author information will not
be reviewed.

\subsubsection{Self-Citations}

If you are citing published papers for which you are an author, refer
to yourself in the third person. In particular, do not use phrases
that reveal your identity (e.g., ``in previous work \cite{langley00}, we
have shown \ldots'').

Do not anonymize citations in the reference section. The only exception are manuscripts that are
not yet published (e.g., under submission). If you choose to refer to
such unpublished manuscripts \cite{anonymous}, anonymized copies have
to be submitted
as Supplementary Material via CMT\@. However, keep in mind that an MLSys
paper should be self contained and should contain sufficient detail
for the reviewers to evaluate the work. In particular, reviewers are
not required to look at the Supplementary Material when writing their
review.

\subsubsection{Camera-Ready Author Information}
\label{final author}

If a paper is accepted, a final camera-ready copy must be prepared.
%
For camera-ready papers, author information should start 0.3~inches below the
bottom rule surrounding the title. The authors' names should appear in 10~point
bold type, in a row, separated by white space, and centered. Author names should
not be broken across lines. Unbolded superscripted numbers, starting 1, should
be used to refer to affiliations.

Affiliations should be numbered in the order of appearance. A single footnote
block of text should be used to list all the affiliations.

Each distinct affiliations should be listed once. If an author has multiple
affiliations, multiple superscripts should be placed after the name, separated
by thin spaces. If the authors would like to highlight equal contribution by
multiple first authors, those authors should have an asterisk placed after their
name in superscript, and the term ``\textsuperscript{*}Equal contribution"
should be placed in the footnote block ahead of the list of affiliations. A
list of corresponding authors and their emails (in the format Full Name
\textless{}email@domain.com\textgreater{}) can follow the list of affiliations.
Ideally only one or two names should be listed.

A sample file with author names is included in the mlsys2024 style file
package. Turn on the \texttt{[accepted]} option to the stylefile to
see the names rendered. All of the guidelines above are implemented
by the \LaTeX\ style file.

\subsection{Abstract}

The paper abstract should begin in the left column, 0.4~inches below the final
address. The heading `Abstract' should be centered, bold, and in 11~point type.
The abstract body should use 10~point type, with a vertical spacing of
11~points, and should be indented 0.25~inches more than normal on left-hand and
right-hand margins. Insert 0.4~inches of blank space after the body. Keep your
abstract brief and self-contained, limiting it to one paragraph and roughly 4--6
sentences. Gross violations will require correction at the camera-ready phase.

\subsection{Partitioning the Text}

You should organize your paper into sections and paragraphs to help
readers place a structure on the material and understand its
contributions.

\subsubsection{Sections and Subsections}

Section headings should be numbered, flush left, and set in 11~pt bold
type with the content words capitalized. Leave 0.25~inches of space
before the heading and 0.15~inches after the heading.

Similarly, subsection headings should be numbered, flush left, and set
in 10~pt bold type with the content words capitalized. Leave
0.2~inches of space before the heading and 0.13~inches afterward.

Finally, subsubsection headings should be numbered, flush left, and
set in 10~pt small caps with the content words capitalized. Leave
0.18~inches of space before the heading and 0.1~inches after the
heading.

Please use no more than three levels of headings.

\subsubsection{Paragraphs and Footnotes}

Within each section or subsection, you should further partition the
paper into paragraphs. Do not indent the first line of a given
paragraph, but insert a blank line between succeeding ones.

You can use footnotes\footnote{Footnotes
should be complete sentences.} to provide readers with additional
information about a topic without interrupting the flow of the paper.
Indicate footnotes with a number in the text where the point is most
relevant. Place the footnote in 9~point type at the bottom of the
column in which it appears. Precede the first footnote in a column
with a horizontal rule of 0.8~inches.\footnote{Multiple footnotes can
appear in each column, in the same order as they appear in the text,
but spread them across columns and pages if possible.}

\subsection{Figures}

You may want to include figures in the paper to illustrate
your approach and results. Such artwork should be centered,
legible, and separated from the text. Lines should be dark and at
least 0.5~points thick for purposes of reproduction, and text should
not appear on a gray background.

Label all distinct components of each figure. If the figure takes the
form of a graph, then give a name for each axis and include a legend
that briefly describes each curve. Do not include a title inside the
figure; instead, the caption should serve this function.

Number figures sequentially, placing the figure number and caption
\emph{after} the graphics, with at least 0.1~inches of space before
the caption and 0.1~inches after it. The figure caption should be set in
9~point type and centered unless it runs two or more lines, in which
case it should be flush left. You may float figures to the top or
bottom of a column, and you may set wide figures across both columns
(use the environment \texttt{figure*} in \LaTeX). Always place
two-column figures at the top or bottom of the page.

\subsection{Algorithms}

If you are using \LaTeX, please use the ``algorithm'' and ``algorithmic''
environments to format pseudocode. These require
the corresponding stylefiles, algorithm.sty and
algorithmic.sty, which are supplied with this package.
Algorithm~\ref{alg:example} shows an example.

\begin{algorithm}[tb]
   \caption{Bubble Sort}
   \label{alg:example}
\begin{algorithmic}
   \STATE {\bfseries Input:} data $x_i$, size $m$
   \REPEAT
   \STATE Initialize $noChange = true$.
   \FOR{$i=1$ {\bfseries to} $m-1$}
   \IF{$x_i > x_{i+1}$}
   \STATE Swap $x_i$ and $x_{i+1}$
   \STATE $noChange = false$
   \ENDIF
   \ENDFOR
   \UNTIL{$noChange$ is $true$}
\end{algorithmic}
\end{algorithm}

\subsection{Tables}

You may also want to include tables that summarize material. Like
figures, these should be centered, legible, and numbered consecutively.
However, place the title \emph{above} the table with at least
0.1~inches of space before the title and the same after it, as in
Table~\ref{sample-table}. The table title should be set in 9~point
type and centered unless it runs two or more lines, in which case it
should be flush left.

% Note use of \abovespace and \belowspace to get reasonable spacing
% above and below tabular lines.

\begin{table}[t]
\caption{Classification accuracies for naive Bayes and flexible
Bayes on various data sets.}
\label{sample-table}
\vskip 0.15in
\begin{center}
\begin{small}
\begin{sc}
\begin{tabular}{lcccr}
\toprule
Data set & Naive & Flexible & Better? \\
\midrule
Breast    & 95.9$\pm$ 0.2& 96.7$\pm$ 0.2& $\surd$ \\
Cleveland & 83.3$\pm$ 0.6& 80.0$\pm$ 0.6& $\times$\\
Glass2    & 61.9$\pm$ 1.4& 83.8$\pm$ 0.7& $\surd$ \\
Credit    & 74.8$\pm$ 0.5& 78.3$\pm$ 0.6&         \\
Horse     & 73.3$\pm$ 0.9& 69.7$\pm$ 1.0& $\times$\\
Meta      & 67.1$\pm$ 0.6& 76.5$\pm$ 0.5& $\surd$ \\
Pima      & 75.1$\pm$ 0.6& 73.9$\pm$ 0.5&         \\
Vehicle   & 44.9$\pm$ 0.6& 61.5$\pm$ 0.4& $\surd$ \\
\bottomrule
\end{tabular}
\end{sc}
\end{small}
\end{center}
\vskip -0.1in
\end{table}

Tables contain textual material, whereas figures contain graphical material.
Specify the contents of each row and column in the table's topmost
row. Again, you may float tables to a column's top or bottom, and set
wide tables across both columns. Place two-column tables at the
top or bottom of the page.

\subsection{Citations and References}

Please use APA reference format regardless of your formatter
or word processor. If you rely on the \LaTeX\/ bibliographic
facility, use \texttt{natbib.sty} and \texttt{mlsys2024.bst}
included in the style-file package to obtain this format.

Citations within the text should include the authors' last names and
year. If the authors' names are included in the sentence, place only
the year in parentheses, for example when referencing Arthur Samuel's
pioneering work \yrcite{Samuel59}. Otherwise place the entire
reference in parentheses with the authors and year separated by a
comma \cite{Samuel59}. List multiple references separated by
semicolons \cite{kearns89,Samuel59,mitchell80}. Use the `et~al.'
construct only for citations with three or more authors or after
listing all authors to a publication in an earlier reference \cite{MachineLearningI}.

Authors should cite their own work in the third person
in the initial version of their paper submitted for blind review.
Please refer to Section~\ref{author info} for detailed instructions on how to
cite your own papers.

Use an unnumbered first-level section heading for the references, and use a
hanging indent style, with the first line of the reference flush against the
left margin and subsequent lines indented by 10 points. The references at the
end of this document give examples for journal articles \cite{Samuel59},
conference publications \cite{langley00}, book chapters \cite{Newell81}, books
\cite{DudaHart2nd}, edited volumes \cite{MachineLearningI}, technical reports
\cite{mitchell80}, and dissertations \cite{kearns89}.

Alphabetize references by the surnames of the first authors, with
single author entries preceding multiple author entries. Order
references for the same authors by year of publication, with the
earliest first. Make sure that each reference includes all relevant
information (e.g., page numbers).

Please put some effort into making references complete, presentable, and
consistent. If using bibtex, please protect capital letters of names and
abbreviations in titles, for example, use \{B\}ayesian or \{L\}ipschitz
in your .bib file.

\subsection{Software and Data}

We strongly encourage the publication of software and data with the
camera-ready version of the paper whenever appropriate. This can be
done by including a URL in the camera-ready copy. However, do not
include URLs that reveal your institution or identity in your
submission for review. Instead, provide an anonymous URL or upload
the material as ``Supplementary Material'' into the CMT reviewing
system. Note that reviewers are not required to look at this material
when writing their review.

% Acknowledgements should only appear in the accepted version.
\section*{Acknowledgements}

\textbf{Do not} include acknowledgements in the initial version of
the paper submitted for blind review.

If a paper is accepted, the final camera-ready version can (and
probably should) include acknowledgements. In this case, please
place such acknowledgements in an unnumbered section at the
end of the paper. Typically, this will include thanks to reviewers
who gave useful comments, to colleagues who contributed to the ideas,
and to funding agencies and corporate sponsors that provided financial
support.

% In the unusual situation where you want a paper to appear in the
% references without citing it in the main text, use \nocite
\nocite{langley00}

\bibliography{example_paper}
\bibliographystyle{mlsys2024}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% SUPPLEMENTAL CONTENT AS APPENDIX AFTER REFERENCES
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\appendix
\section{Please add supplemental material as appendix here}
%
Put anything that you might normally include after the references as an appendix here, {\it not in a separate supplementary file}. Upload your final camera-ready as a single pdf, including all appendices.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\end{document}

% This document was modified from the file originally made available by
% Pat Langley and Andrea Danyluk for ICML-2K. This version was created
% by Iain Murray in 2018. It was modified from a version from Dan Roy in
% 2017, which was based on a version from Lise Getoor and Tobias
% Scheffer, which was slightly modified from the 2010 version by
% Thorsten Joachims & Johannes Fuernkranz, slightly modified from the
% 2009 version by Kiri Wagstaff and Sam Roweis's 2008 version, which is
% slightly modified from Prasad Tadepalli's 2007 version which is a
% lightly changed version of the previous year's version by Andrew
% Moore, which was in turn edited from those of Kristian Kersting and
% Codrina Lauth. Alex Smola contributed to the algorithmic style files.

### 2023-12-07 docs/mlsys2024template/example_paper.tex
add mlsys2024 example

--- 2023-12-07 docs/example_paper.tex
@@
 % Use the following line for the initial blind version submitted for review:
-% \usepackage{mlsys2024}
+\usepackage{mlsys2024}
 
 % If accepted, instead use the following line for the camera-ready submission:
-\usepackage[accepted]{mlsys2024}
+% \usepackage[accepted]{mlsys2024}
 
@@
 % Therefore, a short form for the running title is supplied here:
-\mlsystitlerunning{Step-by-Step: Training IMU-based Gestures with Live Feedback}
+\mlsystitlerunning{Submission and Formatting Instructions for MLSys 2024}
 
@@
 \twocolumn[
-\mlsystitle{Step-by-Step: Training IMU-based Gestures with Live Feedback}
+\mlsystitle{Submission and Formatting Instructions for MLSys 2024}
 
@@
 \begin{mlsysauthorlist}
-\mlsysauthor{Michael Schnebly}{h}
+\mlsysauthor{Aeiau Zzzz}{equal,to}
+\mlsysauthor{Bauiu C.~Yyyy}{equal,to,goo}
+\mlsysauthor{Cieua Vvvvv}{goo}
+\mlsysauthor{Iaesut Saoeu}{ed}
+\mlsysauthor{Fiuea Rrrr}{to}
+\mlsysauthor{Tateu H.~Yasehe}{ed,to,goo}
+\mlsysauthor{Aaoeu Iasoh}{goo}
+\mlsysauthor{Buiui Eueu}{ed}
+\mlsysauthor{Aeuia Zzzz}{ed}
+\mlsysauthor{Bieea C.~Yyyy}{to,goo}
+\mlsysauthor{Teoau Xxxx}{ed}
+\mlsysauthor{Eee Pppp}{ed}
 \end{mlsysauthorlist}
 
-\mlsysaffiliation{h}{School of Engineering and Applied Sciences, Harvard University, Cambridge, MA, USA}
+\mlsysaffiliation{to}{Department of Computation, University of Torontoland, Torontoland, Canada}
+\mlsysaffiliation{goo}{Googol ShallowMind, New London, Michigan, USA}
+\mlsysaffiliation{ed}{School of Computation, University of Edenborrow, Edenborrow, United Kingdom}
 
-\mlsyscorrespondingauthor{Michael Schnebly}{michael\_schnebly@g.harvard.edu}
+\mlsyscorrespondingauthor{Cieua Vvvvv}{c.vvvvv@googol.com}
+\mlsyscorrespondingauthor{Eee Pppp}{ep@eden.co.uk}
 
@@
 \begin{abstract}
-Recognizing user-defined gestures in IMU data unlocks new forms of creativity and accessibility in human-computer interaction.
-However, training gesture recognition models is a difficult task that requires a deep understanding of machine learning.
-We present Step-by-Step, a software tool that allows users to train gesture recognition models with live visual and auditory feedback.
-Step-by-Step is designed to be accessible to users with no machine learning experience, while providing a powerful codebase for advanced users.
-This work makes the following contributions: ...
-% (1) a novel interface for training gesture recognition models with live feedback, (2) a novel interface for visualizing IMU data, and (3) a novel interface for visualizing gesture recognition models.
+This document provides a basic paper template and submission guidelines.
+Abstracts must be a single paragraph, ideally between 4--6 sentences long.
+Gross violations will trigger corrections at the camera-ready phase.
 \end{abstract}
@@
 
-\printAffiliationsAndNotice{}  % leave blank if no need to mention equal contribution
-% \printAffiliationsAndNotice{\mlsysEqualContribution} % otherwise use the standard text.
+%\printAffiliationsAndNotice{}  % leave blank if no need to mention equal contribution
+\printAffiliationsAndNotice{\mlsysEqualContribution} % otherwise use the standard text.
 

### 2023-12-07 docs/latex/paper.tex
reorganized latex

--- 2023-12-07 docs/example_paper.tex
@@
 \begin{abstract}
-Recognizing user-defined gestures in IMU data unlocks new forms of creativity and accessibility in human-computer interaction.
+Recognizing user-defined gestures in inertial measurement unit (IMU) data unlocks new forms of creativity and accessibility in human-computer interaction.
 However, training gesture recognition models is a difficult task that requires a deep understanding of machine learning.
-We present Step-by-Step, a software tool that allows users to train gesture recognition models with live visual and auditory feedback.
+We present Step-by-Step, a software tool that allows users to train gesture recognition models with live audiovisual feedback.
+Step-by-Step uses a simple neural network to learn to recognize and distinguish multiple gestures in IMU timeseries data.
+Users can train the model by performing gestures and receiving live feedback on the model's performance.
 Step-by-Step is designed to be accessible to users with no machine learning experience, while providing a powerful codebase for advanced users.
-This work makes the following contributions: ...
-% (1) a novel interface for training gesture recognition models with live feedback, (2) a novel interface for visualizing IMU data, and (3) a novel interface for visualizing gesture recognition models.
 \end{abstract}

### 2023-12-08 docs/latex/paper.tex

@@
-%%%%%%%% mlsys 2024 EXAMPLE LATEX SUBMISSION FILE %%%%%%%%%%%%%%%%%
-
 \documentclass{article}
-
-% Recommended, but optional, packages for figures and better typesetting:
 \usepackage{microtype}
@@
 \usepackage{booktabs} % for professional tables
-
-% hyperref makes hyperlinks in the resulting PDF.
-% If your build breaks (sometimes temporarily if a hyperlink spans a page)
-% please comment out the following usepackage line and replace
-% \usepackage{mlsys2024} with \usepackage[nohyperref]{mlsys2024} above.
 \usepackage{hyperref}
-
-% Attempt to make hyperref and algorithmic work together better:
 \newcommand{\theHalgorithm}{\arabic{algorithm}}
-
-% Use the following line for the initial blind version submitted for review:
-% \usepackage{mlsys2024}
-
-% If accepted, instead use the following line for the camera-ready submission:
 \usepackage[accepted]{mlsys2024}
 
-% The \mlsystitle you define below is probably too long as a header.
-% Therefore, a short form for the running title is supplied here:
 \mlsystitlerunning{Step-by-Step: Training IMU-based Gestures with Live Feedback}
@@
 \begin{document}
-
 \twocolumn[
 \mlsystitle{Step-by-Step: Training IMU-based Gestures with Live Feedback}
-
-% It is OKAY to include author information, even for blind
-% submissions: the style file will automatically remove it for you
-% unless you've provided the [accepted] option to the mlsys2024
-% package.
-
-% List of affiliations: The first argument should be a (short)
-% identifier you will use later to specify author affiliations
-% Academic affiliations should list Department, University, City, Region, Country
-% Industry affiliations should list Company, City, Region, Country
-
-% You can specify symbols, otherwise they are numbered in order.
-% Ideally, you should not use this facility. Affiliations will be numbered
-% in order of appearance and this is the preferred way.
-\mlsyssetsymbol{equal}{*}
 
@@
 \mlsysaffiliation{h}{School of Engineering and Applied Sciences, Harvard University, Cambridge, MA, USA}
-
 \mlsyscorrespondingauthor{Michael Schnebly}{michael\_schnebly@g.harvard.edu}
 
-% You may provide any keywords that you
-% find helpful for describing your paper; these are used to populate
-% the "keywords" metadata in the PDF but will not be shown in the document
 \mlsyskeywords{Machine Learning, MLSys}
@@
 
-% this must go after the closing bracket ] following \twocolumn[ ...
+\printAffiliationsAndNotice{}
 
-% This command actually creates the footnote in the first column
-% listing the affiliations and the copyright notice.
-% The command takes one argument, which is text to display at the start of the footnote.
-% The \mlsysEqualContribution command is standard text for equal contribution.
-% Remove it (just {}) if you do not need this facility.
+\section{Introduction}
 
-\printAffiliationsAndNotice{}  % leave blank if no need to mention equal contribution
-% \printAffiliationsAndNotice{\mlsysEqualContribution} % otherwise use the standard text.
+\section{Background}
 
-\section{Electronic Submission}
-\label{submission}
+\section{Method}
 
-Submission to MLSys 2024 will be entirely electronic, via a web site
-(not email). Information about the submission process and \LaTeX\ templates
-are available on the conference web site at:
-\begin{center}
-\textbf{\texttt{https://www.mlsys.org/}}
-\end{center}
-
-The guidelines below will be enforced for initial submissions and
-camera-ready copies. Here is a brief summary:
-\begin{itemize}
-\item Submissions must be in PDF\@.
-\item The maximum paper length is \textbf{10 pages} excluding references and appendices.
-    (pages 11 and on must contain only references).
-\item \textbf{Do not include author information or acknowledgements} in your
-    initial submission.
-\item Your paper should be in \textbf{10 point Times font}.
-\item Make sure your PDF file only uses Type-1 fonts.
-\item Place figure captions \emph{under} the figure (and omit titles from inside
-    the graphic file itself). Place table captions \emph{over} the table.
-\item References must include page numbers whenever possible and be as complete
-    as possible. Place multiple citations in chronological order.
-\item Do not alter the style template; in particular, do not compress the paper
-    format by reducing the vertical spaces.
-\item Keep your abstract brief and self-contained, one paragraph and roughly
-    4--6 sentences. Gross violations will require correction at the
-    camera-ready phase. The title should have content words capitalized.
-\item Submissions that significantly diverge from the specified format 
-    will be rejected without review.
-\end{itemize}
-
-\subsection{Submitting Papers}
-
-\textbf{Paper Deadline:} The deadline for paper submission that is
-advertised on the conference website is strict. If your full,
-anonymized, submission does not reach us on time, it will not be
-considered for publication. There is no separate abstract submission.
-
-\textbf{Anonymous Submission:} MLSys uses double-blind review: no identifying
-author information may appear on the title page or in the paper
-itself. Section~\ref{author info} gives further details.
-
-\textbf{Simultaneous Submission:} MLSys will not accept any paper which,
-at the time of submission, is under review for another conference or
-has already been published. This policy also applies to papers that
-overlap substantially in technical content with conference papers
-under review or previously published. MLSys submissions must not be
-submitted to other conferences during MLSys's review period. Authors
-may submit to MLSys substantially different versions of journal papers
-that are currently under review by the journal, but not yet accepted
-at the time of submission. Informal publications, such as technical
-reports or papers in workshop proceedings which do not appear in
-print, do not fall under these restrictions.
-
-\medskip
-
-Authors must provide their manuscripts in \textbf{PDF} format.
-Furthermore, please make sure that files contain only embedded Type-1 fonts
-(e.g.,~using the program \texttt{pdffonts} in linux or using
-File/DocumentProperties/Fonts in Acrobat). Other fonts (like Type-3)
-might come from graphics files imported into the document.
-
-Authors using \textbf{Word} must convert their document to PDF\@. Most
-of the latest versions of Word have the facility to do this
-automatically. Submissions will not be accepted in Word format or any
-format other than PDF\@.
-
-Those who use \textbf{\LaTeX} should avoid including Type-3 fonts.
-Those using \texttt{latex} and \texttt{dvips} may need the following
-two commands:
-
-{\footnotesize
-\begin{verbatim}
-dvips -Ppdf -tletter -G0 -o paper.ps paper.dvi
-ps2pdf paper.ps
-\end{verbatim}}
-It is a zero following the ``-G'', which tells dvips to use
-the config.pdf file. Newer \TeX\ distributions don't always need this
-option.
-
-Using \texttt{pdflatex} rather than \texttt{latex}, often gives better
-results. This program avoids the Type-3 font problem, and supports more
-advanced features in the \texttt{microtype} package.
-
-\textbf{Graphics files} should be a reasonable size, and included from
-an appropriate format. Use vector formats (.eps/.pdf) for plots,
-lossless bitmap formats (.png) for raster graphics with sharp lines, and
-jpeg for photo-like images.
-
-The style file uses the \texttt{hyperref} package to make clickable
-links in documents. If this causes problems for you, add
-\texttt{nohyperref} as one of the options to the \texttt{mlsys2024}
-usepackage statement.
-
-\subsection{Submitting Final Camera-Ready Copy}
-
-The final versions of papers accepted for publication should follow the
-same format and naming convention as initial submissions, except that
-author information (names and affiliations) should be given. See
-Section~\ref{final author} for formatting instructions.
-
-The footnote, ``Preliminary work. Under review by the MLSys. Do not distribute.'' 
-must be modified to ``\textit{Proceedings of the
-$\mathit{5}^{th}$ MLSys Conference},
-Santa Clara, CA, USA, 2024.
-Copyright 2024 by the author(s).''
-
-For those using the \textbf{\LaTeX} style file, this change (and others) is
-handled automatically by simply changing
-$\mathtt{\backslash usepackage\{mlsys2024\}}$ to
-$$\mathtt{\backslash usepackage[accepted]\{mlsys2024\}}$$
-Authors using \textbf{Word} must edit the
-footnote on the first page of the document themselves.
-
-Camera-ready copies should have the title of the paper as running head
-on each page except the first one. The running title consists of a
-single line centered above a horizontal rule which is $1$~point thick.
-The running head should be centered, bold and in $9$~point type. The
-rule should be $10$~points above the main text. For those using the
-\textbf{\LaTeX} style file, the original title is automatically set as running
-head using the \texttt{fancyhdr} package which is included in the MLSys
-2024 style file package. In case that the original title exceeds the
-size restrictions, a shorter form can be supplied by using
-
-\verb|\mlsystitlerunning{...}|
-
-just before $\mathtt{\backslash begin\{document\}}$.
-Authors using \textbf{Word} must edit the header of the document themselves.
-
-\section{Format of the Paper}
-
-All submissions must follow the specified format.
-
-\subsection{Length and Dimensions}
-
-Papers must not exceed ten (10) pages, including all figures, tables,
-but excluding acknowledgements, references and appendices.
-Acknowledgements should be limited to grants and people who contributed to the paper.
-Any submission that exceeds
-this page limit, or that diverges significantly from the specified format,
-will be rejected without review.
-
-The text of the paper should be formatted in two columns, with an
-overall width of 6.75~inches, height of 9.0~inches, and 0.25~inches
-between the columns. The left margin should be 0.75~inches and the top
-margin 1.0~inch (2.54~cm). The right and bottom margins will depend on
-whether you print on US letter or A4 paper, but all final versions
-must be produced for US letter size.
-
-The paper body should be set in 10~point type with a vertical spacing
-of 11~points. Please use Times typeface throughout the text.
-
-\subsection{Title}
-
-The paper title should be set in 14~point bold type and centered
-between two horizontal rules that are 1~point thick, with 1.0~inch
-between the top rule and the top edge of the page. Capitalize the
-first letter of content words and put the rest of the title in lower
-case.
-
-\subsection{Author Information for Submission}
-\label{author info}
-
-MLSys uses double-blind review, so author information must not appear. If
-you are using \LaTeX\/ and the \texttt{mlsys2024.sty} file, use
-\verb+\mlsysauthor{...}+ to specify authors and \verb+\mlsysaffiliation{...}+ to specify affiliations. (Read the TeX code used to produce this document for an example usage.) The author information
-will not be printed unless \texttt{accepted} is passed as an argument to the
-style file.
-Submissions that include the author information will not
-be reviewed.
-
-\subsubsection{Self-Citations}
-
-If you are citing published papers for which you are an author, refer
-to yourself in the third person. In particular, do not use phrases
-that reveal your identity (e.g., ``in previous work \cite{langley00}, we
-have shown \ldots'').
-
-Do not anonymize citations in the reference section. The only exception are manuscripts that are
-not yet published (e.g., under submission). If you choose to refer to
-such unpublished manuscripts \cite{anonymous}, anonymized copies have
-to be submitted
-as Supplementary Material via CMT\@. However, keep in mind that an MLSys
-paper should be self contained and should contain sufficient detail
-for the reviewers to evaluate the work. In particular, reviewers are
-not required to look at the Supplementary Material when writing their
-review.
-
-\subsubsection{Camera-Ready Author Information}
-\label{final author}
-
-If a paper is accepted, a final camera-ready copy must be prepared.
-%
-For camera-ready papers, author information should start 0.3~inches below the
-bottom rule surrounding the title. The authors' names should appear in 10~point
-bold type, in a row, separated by white space, and centered. Author names should
-not be broken across lines. Unbolded superscripted numbers, starting 1, should
-be used to refer to affiliations.
-
-Affiliations should be numbered in the order of appearance. A single footnote
-block of text should be used to list all the affiliations.
-
-Each distinct affiliations should be listed once. If an author has multiple
-affiliations, multiple superscripts should be placed after the name, separated
-by thin spaces. If the authors would like to highlight equal contribution by
-multiple first authors, those authors should have an asterisk placed after their
-name in superscript, and the term ``\textsuperscript{*}Equal contribution"
-should be placed in the footnote block ahead of the list of affiliations. A
-list of corresponding authors and their emails (in the format Full Name
-\textless{}email@domain.com\textgreater{}) can follow the list of affiliations.
-Ideally only one or two names should be listed.
-
-A sample file with author names is included in the mlsys2024 style file
-package. Turn on the \texttt{[accepted]} option to the stylefile to
-see the names rendered. All of the guidelines above are implemented
-by the \LaTeX\ style file.
-
-\subsection{Abstract}
-
-The paper abstract should begin in the left column, 0.4~inches below the final
-address. The heading `Abstract' should be centered, bold, and in 11~point type.
-The abstract body should use 10~point type, with a vertical spacing of
-11~points, and should be indented 0.25~inches more than normal on left-hand and
-right-hand margins. Insert 0.4~inches of blank space after the body. Keep your
-abstract brief and self-contained, limiting it to one paragraph and roughly 4--6
-sentences. Gross violations will require correction at the camera-ready phase.
-
-\subsection{Partitioning the Text}
-
-You should organize your paper into sections and paragraphs to help
-readers place a structure on the material and understand its
-contributions.
-
-\subsubsection{Sections and Subsections}
-
-Section headings should be numbered, flush left, and set in 11~pt bold
-type with the content words capitalized. Leave 0.25~inches of space
-before the heading and 0.15~inches after the heading.
-
-Similarly, subsection headings should be numbered, flush left, and set
-in 10~pt bold type with the content words capitalized. Leave
-0.2~inches of space before the heading and 0.13~inches afterward.
-
-Finally, subsubsection headings should be numbered, flush left, and
-set in 10~pt small caps with the content words capitalized. Leave
-0.18~inches of space before the heading and 0.1~inches after the
-heading.
-
-Please use no more than three levels of headings.
-
-\subsubsection{Paragraphs and Footnotes}
-
-Within each section or subsection, you should further partition the
-paper into paragraphs. Do not indent the first line of a given
-paragraph, but insert a blank line between succeeding ones.
-
-You can use footnotes\footnote{Footnotes
-should be complete sentences.} to provide readers with additional
-information about a topic without interrupting the flow of the paper.
-Indicate footnotes with a number in the text where the point is most
-relevant. Place the footnote in 9~point type at the bottom of the
-column in which it appears. Precede the first footnote in a column
-with a horizontal rule of 0.8~inches.\footnote{Multiple footnotes can
-appear in each column, in the same order as they appear in the text,
-but spread them across columns and pages if possible.}
-
-\subsection{Figures}
-
-You may want to include figures in the paper to illustrate
-your approach and results. Such artwork should be centered,
-legible, and separated from the text. Lines should be dark and at
-least 0.5~points thick for purposes of reproduction, and text should
-not appear on a gray background.
-
-Label all distinct components of each figure. If the figure takes the
-form of a graph, then give a name for each axis and include a legend
-that briefly describes each curve. Do not include a title inside the
-figure; instead, the caption should serve this function.
-
-Number figures sequentially, placing the figure number and caption
-\emph{after} the graphics, with at least 0.1~inches of space before
-the caption and 0.1~inches after it. The figure caption should be set in
-9~point type and centered unless it runs two or more lines, in which
-case it should be flush left. You may float figures to the top or
-bottom of a column, and you may set wide figures across both columns
-(use the environment \texttt{figure*} in \LaTeX). Always place
-two-column figures at the top or bottom of the page.
-
-\subsection{Algorithms}
-
-If you are using \LaTeX, please use the ``algorithm'' and ``algorithmic''
-environments to format pseudocode. These require
-the corresponding stylefiles, algorithm.sty and
-algorithmic.sty, which are supplied with this package.
-Algorithm~\ref{alg:example} shows an example.
-
-\begin{algorithm}[tb]
-   \caption{Bubble Sort}
-   \label{alg:example}
-\begin{algorithmic}
-   \STATE {\bfseries Input:} data $x_i$, size $m$
-   \REPEAT
-   \STATE Initialize $noChange = true$.
-   \FOR{$i=1$ {\bfseries to} $m-1$}
-   \IF{$x_i > x_{i+1}$}
-   \STATE Swap $x_i$ and $x_{i+1}$
-   \STATE $noChange = false$
-   \ENDIF
-   \ENDFOR
-   \UNTIL{$noChange$ is $true$}
-\end{algorithmic}
-\end{algorithm}
-
-\subsection{Tables}
-
-You may also want to include tables that summarize material. Like
-figures, these should be centered, legible, and numbered consecutively.
-However, place the title \emph{above} the table with at least
-0.1~inches of space before the title and the same after it, as in
-Table~\ref{sample-table}. The table title should be set in 9~point
-type and centered unless it runs two or more lines, in which case it
-should be flush left.
-
-% Note use of \abovespace and \belowspace to get reasonable spacing
-% above and below tabular lines.
-
-\begin{table}[t]
-\caption{Classification accuracies for naive Bayes and flexible
-Bayes on various data sets.}
-\label{sample-table}
-\vskip 0.15in
-\begin{center}
-\begin{small}
-\begin{sc}
-\begin{tabular}{lcccr}
-\toprule
-Data set & Naive & Flexible & Better? \\
-\midrule
-Breast    & 95.9$\pm$ 0.2& 96.7$\pm$ 0.2& $\surd$ \\
-Cleveland & 83.3$\pm$ 0.6& 80.0$\pm$ 0.6& $\times$\\
-Glass2    & 61.9$\pm$ 1.4& 83.8$\pm$ 0.7& $\surd$ \\
-Credit    & 74.8$\pm$ 0.5& 78.3$\pm$ 0.6&         \\
-Horse     & 73.3$\pm$ 0.9& 69.7$\pm$ 1.0& $\times$\\
-Meta      & 67.1$\pm$ 0.6& 76.5$\pm$ 0.5& $\surd$ \\
-Pima      & 75.1$\pm$ 0.6& 73.9$\pm$ 0.5&         \\
-Vehicle   & 44.9$\pm$ 0.6& 61.5$\pm$ 0.4& $\surd$ \\
-\bottomrule
-\end{tabular}
-\end{sc}
-\end{small}
-\end{center}
-\vskip -0.1in
-\end{table}
-
-Tables contain textual material, whereas figures contain graphical material.
-Specify the contents of each row and column in the table's topmost
-row. Again, you may float tables to a column's top or bottom, and set
-wide tables across both columns. Place two-column tables at the
-top or bottom of the page.
-
-\subsection{Citations and References}
-
-Please use APA reference format regardless of your formatter
-or word processor. If you rely on the \LaTeX\/ bibliographic
-facility, use \texttt{natbib.sty} and \texttt{mlsys2024.bst}
-included in the style-file package to obtain this format.
-
-Citations within the text should include the authors' last names and
-year. If the authors' names are included in the sentence, place only
-the year in parentheses, for example when referencing Arthur Samuel's
-pioneering work \yrcite{Samuel59}. Otherwise place the entire
-reference in parentheses with the authors and year separated by a
-comma \cite{Samuel59}. List multiple references separated by
-semicolons \cite{kearns89,Samuel59,mitchell80}. Use the `et~al.'
-construct only for citations with three or more authors or after
-listing all authors to a publication in an earlier reference \cite{MachineLearningI}.
-
-Authors should cite their own work in the third person
-in the initial version of their paper submitted for blind review.
-Please refer to Section~\ref{author info} for detailed instructions on how to
-cite your own papers.
-
-Use an unnumbered first-level section heading for the references, and use a
-hanging indent style, with the first line of the reference flush against the
-left margin and subsequent lines indented by 10 points. The references at the
-end of this document give examples for journal articles \cite{Samuel59},
-conference publications \cite{langley00}, book chapters \cite{Newell81}, books
-\cite{DudaHart2nd}, edited volumes \cite{MachineLearningI}, technical reports
-\cite{mitchell80}, and dissertations \cite{kearns89}.
-
-Alphabetize references by the surnames of the first authors, with
-single author entries preceding multiple author entries. Order
-references for the same authors by year of publication, with the
-earliest first. Make sure that each reference includes all relevant
-information (e.g., page numbers).
-
-Please put some effort into making references complete, presentable, and
-consistent. If using bibtex, please protect capital letters of names and
-abbreviations in titles, for example, use \{B\}ayesian or \{L\}ipschitz
-in your .bib file.
-
-\subsection{Software and Data}
-
-We strongly encourage the publication of software and data with the
-camera-ready version of the paper whenever appropriate. This can be
-done by including a URL in the camera-ready copy. However, do not
-include URLs that reveal your institution or identity in your
-submission for review. Instead, provide an anonymous URL or upload
-the material as ``Supplementary Material'' into the CMT reviewing
-system. Note that reviewers are not required to look at this material
-when writing their review.
-
-% Acknowledgements should only appear in the accepted version.
-\section*{Acknowledgements}
-
-\textbf{Do not} include acknowledgements in the initial version of
-the paper submitted for blind review.
-
-If a paper is accepted, the final camera-ready version can (and
-probably should) include acknowledgements. In this case, please
-place such acknowledgements in an unnumbered section at the
-end of the paper. Typically, this will include thanks to reviewers
-who gave useful comments, to colleagues who contributed to the ideas,
-and to funding agencies and corporate sponsors that provided financial
-support.
-
-% In the unusual situation where you want a paper to appear in the
-% references without citing it in the main text, use \nocite
-\nocite{langley00}
+\section{Findings}
 
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% SUPPLEMENTAL CONTENT AS APPENDIX AFTER REFERENCES
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\appendix
-\section{Please add supplemental material as appendix here}
-%
-Put anything that you might normally include after the references as an appendix here, {\it not in a separate supplementary file}.
-Upload your final camera-ready as a single pdf, including all appendices.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
 \end{document}
-
-% This document was modified from the file originally made available by
-% Pat Langley and Andrea Danyluk for ICML-2K. This version was created
-% by Iain Murray in 2018. It was modified from a version from Dan Roy in
-% 2017, which was based on a version from Lise Getoor and Tobias
-% Scheffer, which was slightly modified from the 2010 version by
-% Thorsten Joachims & Johannes Fuernkranz, slightly modified from the
-% 2009 version by Kiri Wagstaff and Sam Roweis's 2008 version, which is
-% slightly modified from Prasad Tadepalli's 2007 version which is a
-% lightly changed version of the previous year's version by Andrew
-% Moore, which was in turn edited from those of Kristian Kersting and
-% Codrina Lauth. Alex Smola contributed to the algorithmic style files.

### 2023-12-11 docs/latex/paper.tex
Update "Methods" Section

@@
 \section{Method}
+% Introductory paragraph summarizing the method's components: hardware, data collection, data labelling, model training, model inference, user interface, and the software architecture.
+% Hardware: what hardware is used, and how it is connected to the computer.
+% Data collection: how the data is collected, what data is collected, and how it is stored.
+% Data labelling: how the data is labelled, what the labels are, and how they are stored.
+% Model structure: what the model is, how it is structured, and how it is stored.
+% Model training: how the model is trained, what the model is, and how it is stored.
+% Model inference: how the model is used to make predictions, and how the predictions are stored.
+% User interface: how the user interacts with the system, and how the system interacts with the user.
+% Software architecture: how the system is structured, and how the components interact with each other.
+
+\subsection{Hardware}
+For measuring body movement, we used the Bosch BNO-055, a 9-axis IMU that includes an accelerometer, gyroscope, and magnetometer.
+The IMU uses I2C communication protocol to stream data to a microcontroller (Espressif ESP8266) which passes that data on to a laptop  (Macbook Pro 2017) via USB for further processing.
+
+Note that the particular hardware components used in this project are not essential to the system.
+They could be replaced with other options so long as the sensor data contains information necessary to recognize and distinguish the gestures of interest and the computer (or microcontroller if using purely embedded approach) is capable of running the software at a rate matching that of data collection.
+% The hardware is shown in Figure \ref{fig:hardware}.
+
+\subsection{Software}
+The software is written in Python and designed to be accessible to users with no machine learning experience while providing a powerful codebase for advanced users.
+The software is structured into three main components: data processing, neural network, and user interface.
+The data processing component handles data collection, labelling, and storage.
+The neural network component handles model structure, training, and inference.
+The user interface component handles user interaction and feedback.
+% The software architecture is shown in Figure \ref{fig:software_architecture}.
+
+\subsection{Data Processing}
+\subsubsection{Collection}
+The IMU data is collected as a timeseries of frames.
+Each frame is a 3D vector representing linear acceleration in each of the three axes: x, y, and z.
+The data is collected at a rate of 100 Hz and can be streamed directly from an IMU, recorded to a file, or loaded from a file.
+
+\subsubsection{Labelling}
+The data is labelled by associating each frame with a gesture.
+A gesture is a sequence of frames that represents a single movement of the body.
+Here, we focus on percussive gestures and assume that a gesture is complete at the local peak in linear acceleration's magnitude.
+This constraint allow us to label gestures using a simple peak-detection algorithm.
+When a local peak meet's user-defined criteria (e.g.
+acceleration is above a certain threshold), the peak frame is labelled as containing a gesture.
+All other frames are labelled as containing no gesture.
+
+\subsection{Neural Network}
+\subsubsection{Structure}
+
+The input to the network is a sequence of frames, each of which is a 3D vector representing linear acceleration in each of the three axes: x, y, and z.
+The output of the network is a vector of probabilities, one for each gesture.
+The network is trained to maximize the probability of the correct gesture and minimize the probabilities of all other gestures.
+The neural network is a shallow, two-branched model.
+One branch represents a recent history of sensor values while the other represents a recent history of model outputs.
+Combining the information from the these two sources, the model is trained to maximize the probability of the correct gesture and minimize the probabilities of all other gestures.
+
+The sensor branch consists of an input layer (short timeseries of recent sensor values) and a 1D convolutional layer (recognizes patterns in the timeseries).
+The memory branch consists of an input layer (short timeseries of recent model outputs) and a max pooling layer (identifies the maximum probability that a gesture has been predicted recently).
+These branches are then concatenated and followed by a fully connected output layer.
+
+\subsubsection{Training}
+
+% The algorithm is described in more detail in the Appendix.
 

### 2023-12-11 docs/latex/paper.tex
Finished Method's Section

@@
 The user interface component handles user interaction and feedback.
+
+To facilitate realtime training and inference, the software is structured as a pipeline with the following processes running in parallel: data processing, model training, and model inference.
+Importantly all processes refer to the same global state, allowing the training data and model to be updated and tested in realtime as new data is collected.
 % The software architecture is shown in Figure \ref{fig:software_architecture}.
@@
 Here, we focus on percussive gestures and assume that a gesture is complete at the local peak in linear acceleration's magnitude.
-This constraint allow us to label gestures using a simple peak-detection algorithm.
+This constraint allow us to achieve realtime, automated labelling of gestures using a simple peak-detection algorithm.
 When a local peak meet's user-defined criteria (e.g.
@@
 \subsubsection{Structure}
-
-The input to the network is a sequence of frames, each of which is a 3D vector representing linear acceleration in each of the three axes: x, y, and z.
-The output of the network is a vector of probabilities, one for each gesture.
-The network is trained to maximize the probability of the correct gesture and minimize the probabilities of all other gestures.
 The neural network is a shallow, two-branched model.
@@
 
-The sensor branch consists of an input layer (short timeseries of recent sensor values) and a 1D convolutional layer (recognizes patterns in the timeseries).
-The memory branch consists of an input layer (short timeseries of recent model outputs) and a max pooling layer (identifies the maximum probability that a gesture has been predicted recently).
+The sensor branch consists of an input layer (short timeseries of recent sensor values) followed by a  a 1D convolutional layer (recognizes patterns in the timeseries).
+The memory branch consists of an input layer (short timeseries of recent model outputs) followed by a max pooling layer (summarizes the timeseries of recent model outputs based on its highest values).
 These branches are then concatenated and followed by a fully connected output layer.
+% The model structure is shown in Figure \ref{fig:model_structure}.
 
-\subsubsection{Training}
+\subsubsection{Realtime Training}
+Training is performed using Tensorflow's default gradient descent optimizer.
+The training data consists of a timeseries of frames, each of which is labelled with a gesture.
+The model is trained on a sliding window of frames where window size is a user-defined parameter.
+For example, window size could be set to 20 frames, equating to 200ms of data at 100Hz.
 
-% The algorithm is described in more detail in the Appendix.
+The window slides forward one frame at a time.
+If the most recent frame in a window is labelled with a gesture, the window is labelled with that gesture.
+Otherwise, the window is labelled with no gesture.
+As new windows are collected, the model is trained on the windows using gradient descent.
+Training is iterative and the model's training data is updated in realtime, allowing the user to receive live feedback on the model's performance.
+
+In principle this training could be done stochastically, training on each frame's window as it is collected.
+However, this would  would require a more sophisticated training algorithm that Tensorflow's built-in gradient descent optimizer (at least given the computational performance of the author's machine).
+For this reason, this proof-of-concept codebase defaults to training on batches of recent windows.
+When training completes, the model is updated, the training data is updated with recent frames, and training begins again.
+
+\subsubsection{Realtime Inference}
+As the model is updated, its performance is tested on the most recent frames using Tensorflow.
+Along with the groundtruth labels, the model's predictions are used to generate live feedback for the user using both auditory and visual cues as discussed in the User Interface section.
+
+\subsection{User Interface}
+The user interface is designed to provide live feedback on the model's performance.
+The user interface consists of two components: a visual component and an auditory component.
+The visual component is a simple GUI that displays the model's predictions and groundtruth labels in realtime.
+The auditory component is a simple synthesizer that plays a tone corresponding to the model's prediction.
+
+Using keyboard inputs, the user can control various aspects of the application. The user can start and stop data collection, start and stop model training, and start and stop model inference.
+% The user interface is shown in Figure \ref{fig:user_interface}.
 

### 2023-12-11 docs/latex/paper.tex
Finish conclusions. Develop introduction/background.

@@
 \section{Introduction}
+For the past 40 years, the keyboard and mouse have been the primary means of human-computer interaction.
+The reliability and precision of these devices is undisputed, but it comes at a cost.
+The limited degrees of freedom available in keyboard and mouse inputs dictate which human body movements can be used to control computers, limiting both accessibility and creativity.
 
-\section{Background}
+The ideal human-computer interface can adapt to a user's needs or preferences.
+It can be used by anyone, regardless of their physical abilities.
+It can be used in any context, regardless of the user's surroundings.
+It can be used to control any device, regardless of its form factor.
+It can be used to control any application, regardless of its purpose.
+
+The challenge in developing such a device is that it must be able to learn, recognize, and distinguish the user's movements.
+This is a difficult task that requires a deep understanding of machine learning.
+
+In this paper, we present Step-by-Step, a software tool that allows users to train gesture recognition models with live audiovisual feedback.
+Step-by-Step uses a simple neural network to learn to recognize and distinguish multiple gestures in IMU timeseries data.
+Users can train the model by performing gestures and receiving live feedback on the model's performance.
+Step-by-Step is designed to be accessible to users with no machine learning experience, while providing a powerful codebase for advanced users.
+
+\subsection{Gesture Recognition}
+Gesture recognition is the process of recognizing and distinguishing gestures in data.
+Gestures are movements of the body that convey information.
+For example, a gesture might be a hand movement that indicates a desire to turn on a light.
+Gesture recognition is a key component of human-computer interaction, allowing users to interact with computers in a natural way.
+
+\subsection{Inertial Measurement Units}
+Inertial measurement units (IMUs) are sensors that measure the movement of the body.
+IMUs are commonly used in wearable devices such as smartwatches and fitness trackers.
+IMUs typically include an accelerometer, gyroscope, and magnetometer, each of which measures in three axes: x, y, and z.
+
+\subsection{Neural Networks}
+Neural networks are a class of machine learning models that are inspired by the structure of the brain.
+Neural networks are composed of layers of neurons, each of which takes in a set of inputs and produces an output.
+The outputs of one layer are used as inputs to the next layer.
+The first layer is called the input layer and the last layer is called the output layer.
+The layers in between are called hidden layers.
+The output layer produces the model's prediction.
 
 \section{Method}
-% Introductory paragraph summarizing the method's components: hardware, data collection, data labelling, model training, model inference, user interface, and the software architecture.
-% Hardware: what hardware is used, and how it is connected to the computer.
-% Data collection: how the data is collected, what data is collected, and how it is stored.
-% Data labelling: how the data is labelled, what the labels are, and how they are stored.
-% Model structure: what the model is, how it is structured, and how it is stored.
-% Model training: how the model is trained, what the model is, and how it is stored.
-% Model inference: how the model is used to make predictions, and how the predictions are stored.
-% User interface: how the user interacts with the system, and how the system interacts with the user.
-% Software architecture: how the system is structured, and how the components interact with each other.
+The system consists of hardware and software components.
+The hardware measures body movement and streams the data to a computer.
+The software processes the data, trains a neural network, and provides live feedback to the user.
+The system is designed to be accessible to users with no machine learning experience while providing a powerful codebase for advanced users.
 
@@
 \subsection{Software}
-The software is written in Python and designed to be accessible to users with no machine learning experience while providing a powerful codebase for advanced users.
-The software is structured into three main components: data processing, neural network, and user interface.
+The software is written in Python and structured into three main components: data processing, neural network, and user interface.
 The data processing component handles data collection, labelling, and storage.
@@
 To facilitate realtime training and inference, the software is structured as a pipeline with the following processes running in parallel: data processing, model training, and model inference.
-Importantly all processes refer to the same global state, allowing the training data and model to be updated and tested in realtime as new data is collected.
+Importantly all processes refer to the same global variables, allowing them to share updates to the model and training data in realtime.
 % The software architecture is shown in Figure \ref{fig:software_architecture}.
@@
 The model is trained on a sliding window of frames where window size is a user-defined parameter.
-For example, window size could be set to 20 frames, equating to 200ms of data at 100Hz.
 
@@
 As the model is updated, its performance is tested on the most recent frames using Tensorflow.
-Along with the groundtruth labels, the model's predictions are used to generate live feedback for the user using both auditory and visual cues as discussed in the User Interface section.
+Ideally, this would be done on a frame-by-frame basis, but computational constraints required the authors to batch the frames and perform inference on batches.
 
-\subsection{User Interface}
-The user interface is designed to provide live feedback on the model's performance.
-The user interface consists of two components: a visual component and an auditory component.
-The visual component is a simple GUI that displays the model's predictions and groundtruth labels in realtime.
-The auditory component is a simple synthesizer that plays a tone corresponding to the model's prediction.
+Along with the groundtruth labels, the model's predictions are used to generate live feedback for the user using both auditory and visual cues.
+The auditory cue is a simple synthesizer that plays a tone corresponding to the model's prediction.
+The visual cue is a simple GUI that displays the IMU timeseries, model's predictions, and groundtruth labels in realtime.
 
-Using keyboard inputs, the user can control various aspects of the application. The user can start and stop data collection, start and stop model training, and start and stop model inference.
-% The user interface is shown in Figure \ref{fig:user_interface}.
+\subsection{User Interface and Control}
+All rendering is done using OpenGL.
+With keyboard inputs, the user can control various aspects of the application.
+The user can start and stop data collection, data labelling, model training, and model inference.
+% The user interface and controls are shown in Figure \ref{fig:user_interface}.
 
-\section{Findings}
+\section{Results}
+After just a few examples of a gesture, the model converges on solution that reliably distinguishes gesture from non-gesture.
+Performance improves as more examples are provided.
+Here, the authors only test the model's ability to recognize a single gesture, but based on prior experience using this model in non-realtime training scenarios, the authors expect the model to be able to distinguish multiple gestures with high accuracy.
+% The model's performance is shown in Figure \ref{fig:performance}.
+
+\section{Next Steps}
+There are many ways in which the system could be improved in terms of hardware, software, and user experience.
+
+The current hardware system uses a microcontroller to stream IMU data to a computer for processing.
+However, given the small size of the model (default settings: ~2kb with ~500 params) and recent advances in embedded machine learning, it should be possible to run the entire system in realtime on a single microcontroller, including labelling, training, and inference.
+This would allow the system to be used in a wider range of applications, including those where a computer is not available, desirable, or practical.
+To this end, the authors are currently exploring the possiblity of using Espressifs ESP32-S3 chip, which includes a neural network accelerator and is capable of running Tensorflow Lite Micro.
+
+The current software is written in Python and uses Tensorflow for training and inference.
+This solution is relatively simple, but is not ideal for realtime applications.
+The authors are currently exploring the possibility of rewriting the software in C++ and using Tensorflow Lite Micro for training and inference.
+This would allow the system to be run on a microcontroller, as described above.
+
+The current user interface is a simple OpenGL-based GUI that displays the IMU timeseries, model's predictions, and groundtruth labels in realtime.
+This solution is relatively barebones and should work on most operating systems, but it is dependent on streaming the data to a personal computer.
+The authors are considering the minimal user-interface necessary to run the system on a microcontroller, including a simple LCD display and a few buttons for control.
+
+\section{Closing Remarks}
+Step-by-Step serves as a proof-of-concept for realtime learning of user-defined gestures in IMU data.
+However, the system is not limited to IMU data and could be extended to other types of bodytracking data such as the 2D and 3D pose estimation.
+The authors hope that this work will inspire others to explore the possibilities of realtime machine learning in human-computer interaction.
+
+\section{Acknowledgements}
+The authors would like the thank Harvard's TinyML course staff for their guidance and support throughout the semester, including: Jason Jabbour, Ike Uchendu, Jessica Quaye, Matthew Stewart, and Vijay Janapa Reddi.
 

### 2023-12-11 docs/latex/paper.tex
First full text draft complete

@@
 
-The ideal human-computer interface can adapt to a user's needs or preferences.
-It can be used by anyone, regardless of their physical abilities.
-It can be used in any context, regardless of the user's surroundings.
-It can be used to control any device, regardless of its form factor.
-It can be used to control any application, regardless of its purpose.
-
+The ideal human-computer interface can adapt to a user's needs or preferences and can be used by anyone, regardless of their physical abilities.
 The challenge in developing such a device is that it must be able to learn, recognize, and distinguish the user's movements.
@@
 
+\section{Background}
 \subsection{Gesture Recognition}
 Gesture recognition is the process of recognizing and distinguishing gestures in data.
-Gestures are movements of the body that convey information.
-For example, a gesture might be a hand movement that indicates a desire to turn on a light.
-Gesture recognition is a key component of human-computer interaction, allowing users to interact with computers in a natural way.
+Gestures are body movement patterns that convey information.
+In a way, keyboards and mice perform gesture recognition by translating body movements into computer inputs.
+However, these devices are limited to a small set of predefined gestures.
+The goal of this work is to allow users to define their own gestures and use them to control computers.
 
 \subsection{Inertial Measurement Units}
-Inertial measurement units (IMUs) are sensors that measure the movement of the body.
-IMUs are commonly used in wearable devices such as smartwatches and fitness trackers.
+Inertial measurement units (IMUs) are one of many types types of sensors that can be used to measure body movement.
 IMUs typically include an accelerometer, gyroscope, and magnetometer, each of which measures in three axes: x, y, and z.
+IMUs are commonly used in wearable devices to track body movements over time.
+In this work, we use IMUs to measure body movement and recognize gestures.
 
 \subsection{Neural Networks}
-Neural networks are a class of machine learning models that are inspired by the structure of the brain.
-Neural networks are composed of layers of neurons, each of which takes in a set of inputs and produces an output.
-The outputs of one layer are used as inputs to the next layer.
-The first layer is called the input layer and the last layer is called the output layer.
-The layers in between are called hidden layers.
-The output layer produces the model's prediction.
+Neural networks are a class of machine learning models that can be used to analyze IMU timeseries and identify patterns such as gestures.
+Neural networks are composed of layers of "neurons", each of which takes in a set of inputs, performs a simple function, and produces an output.
+The outputs of one layer are used as inputs to the next.
+The final layer represents the model's prediction.
+
+Neural networks are trained by providing them with examples of inputs and corresponding correct outputs.
+The model adjusts its parameters to minimize the difference between its predictions and the groundtruth outputs.
+This process is called gradient descent and is performed iteratively until the model converges on a solution.
+Once trained, the model can be used to make predictions on new inputs.
+
+IMUs and neural networks are well-established and have been used together in a wide range of applications for characterizing body movements.
+However, a machine learning system that can be easily trained to recognize user-defined gestures with performance rivaling that of the mouse and keyboard remains elusive.
+The goal of this work is to lay the foundation for such a system.
 
@@
 The current hardware system uses a microcontroller to stream IMU data to a computer for processing.
-However, given the small size of the model (default settings: ~2kb with ~500 params) and recent advances in embedded machine learning, it should be possible to run the entire system in realtime on a single microcontroller, including labelling, training, and inference.
+However, given the small size of the model (default settings: ~2kb with ~500 params) and recent advances in embedded machine learning, it should be possible to run the entire system in realtime on a wireless, wearable microcontroller, including labelling, training, and inference.
 This would allow the system to be used in a wider range of applications, including those where a computer is not available, desirable, or practical.
@@
 Step-by-Step serves as a proof-of-concept for realtime learning of user-defined gestures in IMU data.
-However, the system is not limited to IMU data and could be extended to other types of bodytracking data such as the 2D and 3D pose estimation.
+It is worth noting that the strategies employed here are not limited to IMU data and could be applied to data from other sensors.
 The authors hope that this work will inspire others to explore the possibilities of realtime machine learning in human-computer interaction.

### 2023-12-12 src/hotkeys.py
minor fixes

@@ -83,10 +83,10 @@ class Hotkeys:
         if key == glfw.KEY_3 and action == glfw.PRESS:
             self.LABELING = not self.LABELING
             if self.LABELING:
-                print("4: Starting Labeling...")
+                print("3: Starting Labeling...")
                 self.gesture_data.start_labeling()
             else:
-                print("4: Stopping Labeling...")
+                print("3: Stopping Labeling...")
                 self.gesture_data.stop_labeling()
 
         if key == glfw.KEY_4 and action == glfw.PRESS:
@@ -97,7 +97,7 @@ class Hotkeys:
             else:
                 self.nn_model.stop_inference()
                 self.nn_plot.lines[0].stop()
-            print("5: NN_INFERENCE " + str(self.NN_INFERENCE))
+            print("4: NN_INFERENCE " + str(self.NN_INFERENCE))
 
         if key == glfw.KEY_5 and action == glfw.PRESS:
             self.NN_TRAINING = not self.NN_TRAINING
@@ -105,7 +105,7 @@ class Hotkeys:
                 self.nn_model.start_training()
             else:
                 self.nn_model.stop_training()
-            print("6: NN_TRAINING " + str(self.NN_TRAINING))
+            print("5: NN_TRAINING " + str(self.NN_TRAINING))
 
 
 

### 2023-12-12 src/neuralnet.py

@@ -30,7 +30,7 @@ class NeuralNetData:
         self.output_labels = np.zeros((n_samples, n_labels), dtype=np.float32)
         self.output_results = np.full((n_samples), 0.5, dtype=np.float32)
 
-        self.high_beat = sa.WaveObject.from_wave_file("sounds/metronome_hi.wav")
+        self.high_beat = sa.WaveObject.from_wave_file("assets/sounds/metronome_hi.wav")
     
     def update(self, input_data_window, input_memory_window, output_labels):
         self.batch_count += 1

### 2023-12-12 docs/latex/paper.tex
final paper commit

@@
 
+\begin{figure}[h]
+    \includegraphics[width=0.5\textwidth]{../figures/model_diagram.jpg}
+    \caption{Neural Network Architecture.
+The sensor branch (left) receives 20 frames (200 ms at 100 Hz) of 3-channel (x, y, z) IMU data.
+The sensor branch feeds into a 1D-convolutional layer with 8 filters, each also 20 frames in width.
+The memory branch (right) receives 15 frames of 1-hot encoded prediction data.
+The two branches are concatenated and fed into a fully connected output layer to predict the currently active gesture.}
+    \label{fig:your_label}
+\end{figure}
+
 \subsection{Data Processing}
@@
 All other frames are labelled as containing no gesture.
+
+\begin{figure*}[h]
+    \centering
+    \includegraphics[width=0.9\textwidth]{../figures/realtime_training.jpg}
+    \caption{Live Learning Feedback: The user performs a gesture and receives live feedback on the model's performance.
+Bottom Panel: 10 seconds of training data.
+Acceleration components (x, y, z) are shown in color; magnitude is shown in grey.
+Each magnitude peak is marked by a thin vertical line, indicating it has been labelled as a gesture in the training data.
+Top Panel: 10 seconds of inference data.
+At t = 0 seconds the neural network is completely untrained.
+Naive to what is and is not a gesture, it predicts gesture and non-gesture with equal probability (0.5).
+As the model trains on non-gesture date, it begins to predict non-gesture with higher probability.
+At t = 5 seconds, the model has converged on a solution that reliably distinguishes non-gesture.
+At t = 10 seconds, the model has converged on a solution that reliably distinguishes gestures.
+Given more training data and time to learn, the model's performance would continue to improve with gessture and non-gesture probabilities approaching 1 and 0, respectively.
+See supplementary video for live demo in the github repo.}
+    \label{fig:realtime_training}
+\end{figure*}
 
@@
 Training is iterative and the model's training data is updated in realtime, allowing the user to receive live feedback on the model's performance.
+This process is illustated in Figure \ref{fig:realtime_training}.
 
@@
 The visual cue is a simple GUI that displays the IMU timeseries, model's predictions, and groundtruth labels in realtime.
+The visual elements are illustated in Figure \ref{fig:realtime_training}.
 
@@
 Here, the authors only test the model's ability to recognize a single gesture, but based on prior experience using this model in non-realtime training scenarios, the authors expect the model to be able to distinguish multiple gestures with high accuracy.
+A representative example of the model learning to recognize a single gesture in just 10 seconds is shown in Figure \ref{fig:realtime_training}.
 % The model's performance is shown in Figure \ref{fig:performance}.
@@
 
-\bibliography{example_paper}
-\bibliographystyle{mlsys2024}
+% \bibliography{example_paper}
+% \bibliographystyle{mlsys2024}
 

### 2023-12-12 main.py
removed extra packages - reset pip requiremnts.txt

@@ -1,9 +1,6 @@
 import glfw
 from OpenGL.GL import *
 
-import imgui
-from imgui.integrations.glfw import GlfwRenderer
-
 from src.imu import IMUData
 from src.stream import IMUStream
 from src.plot import IMUPlot, EventPlot, NNPlot

### 2023-12-12 src/neuralnet.py

@@ -3,10 +3,7 @@ from keras.models import Model
 from keras.losses import CategoricalCrossentropy
 from keras.optimizers import Adam
 from keras.utils import plot_model
-# from sklearn.metrics import confusion_matrix
-import matplotlib.pyplot as plt
 import numpy as np
-# import coremltools as ct
 import time
 import threading
 

### 2023-12-12 src/plot.py

@@ -1,7 +1,4 @@
-import math
 import numpy as np
-import glm
-from scipy.spatial.transform import Rotation as R
 from OpenGL.arrays import vbo
 
 

### 2023-12-12 src/render.py

@@ -1,10 +1,5 @@
-import numpy as np
 from OpenGL.GL import *
-from OpenGL.GL.shaders import compileProgram, compileShader
-from OpenGL.arrays import vbo
-import glfw
-import glm
-import imgui
+from OpenGL.GL.shaders import compileProgram
 
 
 class EventRenderer:

### 2023-12-13 main.py
New Username Test

@@ -1,3 +1,4 @@
+# new username test
 import glfw
 from OpenGL.GL import *
 

### 2024-02-21 docs/latex-nime/nime-template.tex
edited paper for nime

--- 2023-12-12 docs/latex/paper.tex
@@
-\documentclass{article}
-\usepackage{microtype}
-\usepackage{graphicx}
-\usepackage{subfigure}
-\usepackage{booktabs} % for professional tables
-\usepackage{hyperref}
-\newcommand{\theHalgorithm}{\arabic{algorithm}}
-\usepackage[accepted]{mlsys2024}
+\documentclass{nime-alternate} % Uncomment when publishing final version
 
-\mlsystitlerunning{Step-by-Step: Training IMU-based Gestures with Live Feedback}
+% Uncomment only one of the ones below
+% \usepackage{anonymize} 		   %Uncomment this line to publish
+\usepackage[blind]{anonymize}%Uncomment this line for blind review
+
+% Package that enables the use of accents and non 
+% standard characters
+\usepackage[utf8]{inputenc}
 
 \begin{document}
-\twocolumn[
-\mlsystitle{Step-by-Step: Training IMU-based Gestures with Live Feedback}
 
-\begin{mlsysauthorlist}
-\mlsysauthor{Michael Schnebly}{h}
-\end{mlsysauthorlist}
+\conferenceinfo{NIME'24,}{4--6 September, Utrecht, The Netherlands.}
+\title{Step-by-Step: Training IMU-based Gestures with Live Feedback}
 
-\mlsysaffiliation{h}{School of Engineering and Applied Sciences, Harvard University, Cambridge, MA, USA}
-\mlsyscorrespondingauthor{Michael Schnebly}{michael\_schnebly@g.harvard.edu}
+\label{key}
 
-\mlsyskeywords{Machine Learning, MLSys}
+\numberofauthors{1}
 
-\vskip 0.3in
+\author{
+% 1st. author
+\alignauthor
+\anonymize{Michael Schnebly}\\
+       \affaddr{\anonymize{School of Engineering and Applied Sciences}}\\
+       \affaddr{\anonymize{Harvard University}}\\
+       \affaddr{\anonymize{Cambridge, MA, USA}}\\
+       \email{\anonymize{michael\_schnebly@g.harvard.edu}}
+}
+% For your initial submission you MUST ANONYMIZE the authors.
 
+\maketitle
 \begin{abstract}
-Recognizing user-defined gestures in inertial measurement unit (IMU) data unlocks new forms of creativity and accessibility in human-computer interaction.
-However, training gesture recognition models is a difficult task that requires a deep understanding of machine learning.
-We present Step-by-Step, a software tool that allows users to train gesture recognition models with live audiovisual feedback.
-Step-by-Step uses a simple neural network to learn to recognize and distinguish multiple gestures in IMU timeseries data.
-Users can train the model by performing gestures and receiving live feedback on the model's performance.
-Step-by-Step is designed to be accessible to users with no machine learning experience, while providing a powerful codebase for advanced users.
+This document serves as an application to NIME's 2024 Student Consortium.
+Within it, I present my interest in joining the NIME research community, as well as my preliminary results for \textit{Step-by-Step}, a software tool that allows users to train gesture recognition models with live audiovisual feedback.
+\textit{Step-by-Step} uses a simple neural network to learn to recognize and distinguish multiple gestures in inertial measurement unit (IMU) timeseries data.
+Users train the model by repeatedly performing a target gesture while receiving live feedback on the model’s ability to recognize that gesture.
+Once satisfactory results are achieved, the user can save the model and deploy it for realtime musical applications, as demonstrated in supplementary live performance videos.
+An anonymized code repository is available at \url{https://anonymous.4open.science/r/step-by-step-7506/}
 \end{abstract} 
-]
+\keywords{Gesture Recognition, IMU, Real-time, Neural Network, Training, Inference, Live Feedback}
+\ccsdesc[300]{Applied computing~Sound and music computing}
+\ccsdesc[300]{Human-centered computing~Gestural input}
+\ccsdesc[300]{Machine learning~Neural networks}
 
-\printAffiliationsAndNotice{}
+% this line creates the CCS Concepts section.
+\printccsdesc
+
+\section{Expected Outcomes}
 
 \section{Introduction}
-For the past 40 years, the keyboard and mouse have been the primary means of human-computer interaction.
-The reliability and precision of these devices is undisputed, but it comes at a cost.
-The limited degrees of freedom available in keyboard and mouse inputs dictate which human body movements can be used to control computers, limiting both accessibility and creativity.
+The broad range of possible mappings between body movement and sound afforded by gesture-based digital musical instruments presents the musician with a creative challenge: Which body movements?
+Which sounds?
+And, what is the mapping between them \cite{mitchell:mapping}?
 
-The ideal human-computer interface can adapt to a user's needs or preferences and can be used by anyone, regardless of their physical abilities.
-The challenge in developing such a device is that it must be able to learn, recognize, and distinguish the user's movements.
-This is a difficult task that requires a deep understanding of machine learning.
+A common approach to this challenge are \textit{learned} mappings, for which the musician records themselves demonstrating a gesture, then trains a machine learning model to recognize and distinguish the target gesture from other body movements.
+Indeed, the NIME community has produced several software tools that facilitate this process \cite{bullock:mllib,fiebrink:wekinator, gillian:toolkit, caramiaux:nimeml, mcpherson:pipeline}.
 
-In this paper, we present Step-by-Step, a software tool that allows users to train gesture recognition models with live audiovisual feedback.
-Step-by-Step uses a simple neural network to learn to recognize and distinguish multiple gestures in IMU timeseries data.
-Users can train the model by performing gestures and receiving live feedback on the model's performance.
-Step-by-Step is designed to be accessible to users with no machine learning experience, while providing a powerful codebase for advanced users.
+These interactive machine learning tools are typically designed for offline training, wherein the musician records training data, trains the model, then tests it.
+With no knowledge of how the model will perform until after training, the musician must iterate, repeatedly recording, training, and testing until a satisfactory result is achieved.
+This opaque, iterative process can be a hindrance to creative flow, especially in live performance settings where improvising novel gesture-sound mappings might be desirable.
+
+Ideally, one would be able to train a model in realtime, receiving live feedback on the model's performance as it learns each time a gesture is performed.
+Receving model performance reports on a per-gesture basis, the musician can build intuition for what the model has learned (e.g.
+which variations on the gesture are recognized and whicvh are not), adjust their gesture demonstrations accordingly, and arrive at a satisfactory result more efficiently.
+
+This is the motivation behind \textit{Step-by-Step}, a software tool that allows users to train gesture recognition models with live audiovisual feedback.
+Here we present the system's hardware and software components, the method by which they are integrated, and preliminary results demonstrating the system's ability to learn to recognize and distinguish gestures in realtime.
 
@@
 Gestures are body movement patterns that convey information.
-In a way, keyboards and mice perform gesture recognition by translating body movements into computer inputs.
-However, these devices are limited to a small set of predefined gestures.
-The goal of this work is to allow users to define their own gestures and use them to control computers.
+In the context of digital musical instruments, gestures are often used to control sound.
+For example, a musician might use a gesture to control the pitch of a note, the volume of a sound, or the timbre of an instrument.
 
@@
 \subsection{Neural Networks}
-Neural networks are a class of machine learning models that can be used to analyze IMU timeseries and identify patterns such as gestures.
-Neural networks are composed of layers of "neurons", each of which takes in a set of inputs, performs a simple function, and produces an output.
-The outputs of one layer are used as inputs to the next.
-The final layer represents the model's prediction.
-
-Neural networks are trained by providing them with examples of inputs and corresponding correct outputs.
-The model adjusts its parameters to minimize the difference between its predictions and the groundtruth outputs.
+Neural networks are one of many classes of machine learning models that can be used to analyze IMU timeseries and identify patterns such as gestures.
+In \textit{supervised learning}, neural networks are trained by providing them with examples of inputs and corresponding correct output labels.
+During training, the model adjusts its parameters to minimize the difference between its predictions and the groundtruth outputs.
 This process is called gradient descent and is performed iteratively until the model converges on a solution.
-Once trained, the model can be used to make predictions on new inputs.
-
-IMUs and neural networks are well-established and have been used together in a wide range of applications for characterizing body movements.
-However, a machine learning system that can be easily trained to recognize user-defined gestures with performance rivaling that of the mouse and keyboard remains elusive.
-The goal of this work is to lay the foundation for such a system.
+Once trained, the model can be used to make predictions on novel inputs.
+Here, we perform gradient descent in parallel to the recording and automated labelling of training data, allowing the model to learn in real-time as new examples are provided.
 
@@
 The two branches are concatenated and fed into a fully connected output layer to predict the currently active gesture.}
-    \label{fig:your_label}
+    \label{fig:neural_network}
 \end{figure}
@@
     \centering
-    \includegraphics[width=0.9\textwidth]{../figures/realtime_training.jpg}
+    \includegraphics[width=1.0\textwidth]{../figures/realtime_training.jpg}
     \caption{Live Learning Feedback: The user performs a gesture and receives live feedback on the model's performance.
@@
 The current hardware system uses a microcontroller to stream IMU data to a computer for processing.
-However, given the small size of the model (default settings: ~2kb with ~500 params) and recent advances in embedded machine learning, it should be possible to run the entire system in realtime on a wireless, wearable microcontroller, including labelling, training, and inference.
+However, given the small size of the model (default settings: ~2kb with ~500 parameters) and recent advances in embedded machine learning, it should be possible to run the entire system in realtime on a wireless, wearable microcontroller, including labelling, training, and inference.
 This would allow the system to be used in a wider range of applications, including those where a computer is not available, desirable, or practical.
-To this end, the authors are currently exploring the possiblity of using Espressifs ESP32-S3 chip, which includes a neural network accelerator and is capable of running Tensorflow Lite Micro.
 
@@
 
-\section{Closing Remarks}
+\section{Conclusion}
 Step-by-Step serves as a proof-of-concept for realtime learning of user-defined gestures in IMU data.
@@
 
-\section{Acknowledgements}
-The authors would like the thank Harvard's TinyML course staff for their guidance and support throughout the semester, including: Jason Jabbour, Ike Uchendu, Jessica Quaye, Matthew Stewart, and Vijay Janapa Reddi.
+% The following two commands are all you need in the
+% initial runs of your .tex file to
+% produce the bibliography for the citations in your paper.
+\bibliographystyle{abbrv}
+\bibliography{nime-references}
 
-% \bibliography{example_paper}
-% \bibliographystyle{mlsys2024}
+%%% Place this command where you want to balance the columns on the last page. 
+%\balancecolumns
 
+% That's all folks!
 \end{document}

### 2024-02-22 docs/latex-nime/nime-template.tex
minor paper edits - flipped .gif

@@
 \usepackage[utf8]{inputenc}
+\usepackage{hyperref}
 
@@
 Users train the model by repeatedly performing a target gesture while receiving live feedback on the model’s ability to recognize that gesture.
-Once satisfactory results are achieved, the user can save the model and deploy it for realtime musical applications, as demonstrated in supplementary live performance videos.
-An anonymized code repository is available at \url{https://anonymous.4open.science/r/step-by-step-7506/}
+Once satisfactory results are achieved, the user can save the model and deploy it for realtime musical applications.
+The training process is demonstrated in a supplementary video and an anonymized code repository is available at: \href{https://anonymous.4open.science/r/step-by-step-61D8/}{\textbf{https://anonymous.4open.science/r/step-by-step-61D8/}}
+
 \end{abstract} 
@@
 
-\section{Expected Outcomes}
-
 \section{Introduction}
@@
 A common approach to this challenge are \textit{learned} mappings, for which the musician records themselves demonstrating a gesture, then trains a machine learning model to recognize and distinguish the target gesture from other body movements.
-Indeed, the NIME community has produced several software tools that facilitate this process \cite{bullock:mllib,fiebrink:wekinator, gillian:toolkit, caramiaux:nimeml, mcpherson:pipeline}.
+Indeed, the NIME community has prod-\\\\-uced a number of software tools that facilitate this process \cite{bullock:mllib,fiebrink:wekinator, gillian:toolkit, caramiaux:nimeml, mcpherson:pipeline}.
 
@@
 With no knowledge of how the model will perform until after training, the musician must iterate, repeatedly recording, training, and testing until a satisfactory result is achieved.
-This opaque, iterative process can be a hindrance to creative flow, especially in live performance settings where improvising novel gesture-sound mappings might be desirable.
+This iterative process with delayed feedback can be a hindrance to creative flow, especially in live performance settings where improvising novel gesture-sound mappings might be desirable.
 
@@
 Receving model performance reports on a per-gesture basis, the musician can build intuition for what the model has learned (e.g.
-which variations on the gesture are recognized and whicvh are not), adjust their gesture demonstrations accordingly, and arrive at a satisfactory result more efficiently.
+which variations on the gesture are recognized and which are not), adjust their gesture demonstrations accordingly, and arrive at a satisfactory result more efficiently.
 
@@
 \subsection{Gesture Recognition}
+Gestures are body movement patterns that convey information.
 Gesture recognition is the process of recognizing and distinguishing gestures in data.
-Gestures are body movement patterns that convey information.
 In the context of digital musical instruments, gestures are often used to control sound.
@@
 IMUs are commonly used in wearable devices to track body movements over time.
-In this work, we use IMUs to measure body movement and recognize gestures.
+In this work, we use IMUs to measure body movement and recognize gestures.\\\\
 
@@
 Neural networks are one of many classes of machine learning models that can be used to analyze IMU timeseries and identify patterns such as gestures.
-In \textit{supervised learning}, neural networks are trained by providing them with examples of inputs and corresponding correct output labels.
-During training, the model adjusts its parameters to minimize the difference between its predictions and the groundtruth outputs.
+In \textit{supervised learning}, neural networks are trained on example inputs paired with corresponding outputs.
+During training, the model adjusts its parameters to minimize the difference between its predictions about what the outputs should be and the groundtruth outputs it was provided with.
 This process is called gradient descent and is performed iteratively until the model converges on a solution.
@@
 The software processes the data, trains a neural network, and provides live feedback to the user.
-The system is designed to be accessible to users with no machine learning experience while providing a powerful codebase for advanced users.
 
@@
 When a local peak meet's user-defined criteria (e.g.
-acceleration is above a certain threshold), the peak frame is labelled as containing a gesture.
+acceleration is above a certain threshold), the peak frame is labelled as containing the current target gesture.
 All other frames are labelled as containing no gesture.
@@
     \centering
-    \includegraphics[width=1.0\textwidth]{../figures/realtime_training.jpg}
+    \includegraphics[width=0.98\textwidth]{../figures/realtime_training.jpg}
     \caption{Live Learning Feedback: The user performs a gesture and receives live feedback on the model's performance.
@@
 Acceleration components (x, y, z) are shown in color; magnitude is shown in grey.
-Each magnitude peak is marked by a thin vertical line, indicating it has been labelled as a gesture in the training data.
+Each magnitude peak is marked by a thin vertical line, indicating it has been automatically recognized and labelled as a gesture in the training data.
 Top Panel: 10 seconds of inference data.
@@
 Naive to what is and is not a gesture, it predicts gesture and non-gesture with equal probability (0.5).
-As the model trains on non-gesture date, it begins to predict non-gesture with higher probability.
+As the model trains on non-gesture data, it begins to predict non-gesture with higher probability.
 At t = 5 seconds, the model has converged on a solution that reliably distinguishes non-gesture.
 At t = 10 seconds, the model has converged on a solution that reliably distinguishes gestures.
-Given more training data and time to learn, the model's performance would continue to improve with gessture and non-gesture probabilities approaching 1 and 0, respectively.
-See supplementary video for live demo in the github repo.}
+Given more training data and time to learn, the model's performance with continue to improve with gessture and non-gesture probabilities approaching 1 and 0 respectively.
+The model then can be trained to distinguish additional unique gestures (at least 4 in the author's experience).
+See supplementary videos for demonstrations.}
     \label{fig:realtime_training}
@@
 The sensor branch consists of an input layer (short timeseries of recent sensor values) followed by a  a 1D convolutional layer (recognizes patterns in the timeseries).
-The memory branch consists of an input layer (short timeseries of recent model outputs) followed by a max pooling layer (summarizes the timeseries of recent model outputs based on its highest values).
+The memory branch consists of an input layer (short timeseries of recent model output values) followed by a max pooling layer (summarizes the timeseries of recent model outputs based on its highest values).
 These branches are then concatenated and followed by a fully connected output layer.
-% The model structure is shown in Figure \ref{fig:model_structure}.
+Model architecture is shown in Figure \ref{fig:neural_network}.
+
+Depending on the context, the memory branch may not be necessary.
+In the context of percussive gestures, the authors have found that this self-reference helps the model to learn that gestures typically do not occur in rapid succession.
+% The model structure is shown in Figure \ref{fig:neural_network}.
 
@@
 In principle this training could be done stochastically, training on each frame's window as it is collected.
-However, this would  would require a more sophisticated training algorithm that Tensorflow's built-in gradient descent optimizer (at least given the computational performance of the author's machine).
+However, this would require a more sophisticated training algorithm than Tensorflow's built-in gradient descent optimizer (at least given the computational performance of the author's machine).
 For this reason, this proof-of-concept codebase defaults to training on batches of recent windows.
@@
 Step-by-Step serves as a proof-of-concept for realtime learning of user-defined gestures in IMU data.
+While currently limitted to percussive gestures (MIDI note on/off), a similar scheme could be used to do realtime learning of continuous mappings (MIDI CC messages) as well.
 It is worth noting that the strategies employed here are not limited to IMU data and could be applied to data from other sensors.
-The authors hope that this work will inspire others to explore the possibilities of realtime machine learning in human-computer interaction.
+
+\section{Expected Outcomes}
+In applying for the NIME 2024 Student Consortium, I hope to demo \textit{Step-by-Step} as a preliminary work, receive feedback from the NIME community, connect with other researchers, and experience NIME gesture-sound mapping demonstrations in-person (there is only so much one can glean from reading a paper).
+Given that NIME-related research is uncommon at my institution, the opportunity to connect with other researchers in the NIME community is particularly exciting.
+During background research for this submission, I discovered many NIME papers that are relevant to this work, and I am eager to connect with the authors of those papers and others who are interested in similar topics.
+
+\section{Supplementary Videos}
+\textit{Please find supplementary videos in the supplementary materials section of the conference application.}
 
